---
title: `r Biocpkg("rols"): an R interface to the Ontology Lookup Service
output:
  BiocStyle::html_document:
     toc: true
     toc_depth: 1
vignette: >
  %\VignetteIndexEntry{The rols interface to the Ontology Lookup Service}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Infrastructure, Bioinformatics, ontology, data}
  %\VignettePackage{rols}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("rols")`<br />
**Authors**: `r packageDescription("rols")[["Author"]]`<br />
**Modified:** `r file.info("rols.Rmd")$mtime`<br />
**Compiled**: `r date()`


```{r env, echo=FALSE}
library("xtable")
suppressPackageStartupMessages(library("GO.db"))
suppressPackageStartupMessages(library("rols"))
nonto <- length(ontologyNames())
```

# Introduction

The [Ontology Lookup Service](http://www.ebi.ac.uk/ontology-lookup/)
(OLS)
[[1](http://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-7-97),
[2](http://nar.oxfordjournals.org/content/36/suppl_2/W372)] is a
spin-off of the PRoteomics IDEntifications database (PRIDE) service,
located at the EBI. OLS provides a unified interface to over hundred
ontologies.


`r Biocpkg("rols")` relies on the `r CRANpkg("httr")` to query the
REST interface at the EBI and accesss the OLS.

# Brief \Biocpkg{rols} overview

## Ontologies

There are `r nonto` ontologies available in the OLS, listed in the
table below. Their name is to be use to defined which ontology to
query.


```{r ontTable, results='asis', echo = FALSE}
tab <- ontologies()
o <- order(tab$Name)
print(xtable(tab[o,],
             align=c("l", "l", "l"),
             label="tab:onto",
             caption='Available ontologies.'),
      tabular.environment="longtable",
      floating=FALSE,
      include.rownames=FALSE)
```

## Interface

The table below summarised the common interface available for the
`r nonto` ontologies.  More information is provided in the respective
manual pages.

```{r funTables, echo = FALSE}
funTab <- data.frame(rbind(
  c("olsVersion", "Returns the OLS version"),
  c("ontologies", "Returns all available ontologies"),
  c("ontologyNames", "Returns all ontologyNames"), 
  c("ontologyLoadDate", "Returns the ontology load date"), 
  c("isIdObsolete", "Is the ontology id obsolete"), 
  c("term", "Returns the term of a given identifier"),
  c("termMetadata", "Retuns an identifier's metadata"),
  c("termXrefs", "Returns the idenifier's ontology cross references"),
  c("rootId", "Retuns the root identifiers of an ontology"), 
  c("allIds", "Returns all identifiers and terms of an ontology"), 
  c("olsQuery", "Returns matching identifiers"), 
  c("parents", "Returns the parent(s) of a term."),
  c("childrenRelations", "Returns the children relation type(s).")
))
colnames(funTab) <- c("Function", "Description")
print(xtable(funTab,
             align=c("l", "l", "l"),
             label="tab:fun",
             caption='Functions available to query the ontologies.'),
      include.rownames=FALSE)
```

# Use case

A researcher might be interested in the trans-Golgi network and
interested in knowing in which ontologies his favourite organelle is
referenced. This can be done by querying all ontologies with a
relevant pattern. The code below describes how to achieve this.


```{r tgnquery, eval = TRUE}
library("rols")
alltgns <- olsQuery("trans-golgi network")
```

As shown below, `r length(unique(sapply(strsplit(names(alltgns), ":"), "[", 1)))`
different ontologies have matched the query string.

```{r tgnqueryShow}
alltgns
allonts <- sapply(strsplit(alltgns, ":"), "[", 1)
onto.tab <- table(allonts)
onto.tab
```

The description of the `r length(onto.tab)` ontologies of interest can then be used to subset the ontology description:

```{r tgnqueryTable}
ontologies()[names(onto.tab),]
```

To restrict the search to a specific ontology of interest, one can
specify the ontolgy name as a parameter to `olsQuery`.


```{r gotgnquery}
gotgns <- olsQuery("trans-golgi network", "GO")
gotgns
```

Details about relevant terms can be retrieved with the `term` and
`termMetadata` functions. This functionality provides on-line access
to the same data that is available in the `r Biocpkg("GO.db")`, and
can be extended to any of the `r nonto` available ontologies.

```{r godetails}
term("GO:0005802", "GO")
mtd <- termMetadata("GO:0005802", "GO")
names(mtd)
mtd
strwrap(mtd["comment"])
strwrap(mtd["definition"])
```

Below, we execute the same query using the \Biocpkg{GO.db} package.

```{r go.db}
GOTERM[["GO:0005802"]]
```

# On-line vs. off-line data

It is possible to observe different results with `r Biocpkg("rols")`
and ` r Biocpkg("GO.db")`, as a result of the different ways they
access the data. `r Biocpkg("rols")` or ` r Biocpkg("biomaRt")`
perform direct online queries, while `r Biocpkg("GO.db")` and other
annotation packages use database snapshot that are updated every
release.

Both approaches have advantages. While online queries allow to obtain
the latest up-to-date information, such approaches rely on network
availability and quality. If reproducibility is a major issue, the
version of the database to be queried can easily be controlled with
off-line approaches. In the case of `r Biocpkg("rols")`, altough the
load date of a specific ontology can be queried with `olsVersion`, it
is not possible to query a specific version of an ontology.

# Changes in `r Biocpkg("rols")` 2.0



# Session information

```{r si, echo=FALSE}
print(sessionInfo(), locale = FALSE)
```
